# (APPENDIX) Appendix {-} 

# Notation

$i$: Index for learner. The count starts from 1.

$j$: Index for practice item. The cardinality of the item id is $J$.  The count starts from 1.

$k,m,n$: States/levels of the latent mastery, usually $m<n$. The count starts from 0.

$t$: Index for practice sequence. The cardinality of the sequence id is $T$.

$r$: States/levels of the observed response. The count starts from 0.

$z$: The type of a learner

$A$: The assigment function that maps the items to the practice sequence. $A_t=j$ means assign item $j$ to sequence $t$. $\mathbf{A}_{t_1,t_2}$ is mapping from sequence position $t_1$ to $t_2$. 

$H$: The exit decision . Binary variable. If the learner stops at sequence $t$,$H_t=1$; otherwise, $H_t=0$. $\mathbf{H}_{t_1,t_2}$ is the joint exit decision $(H_{t_1},\dots,H_{t_2})$.

$E$: The effort decision. Binary variable. If the learner exerts effort at sequence $t$, $E_t=1$; otherwise, $E_t=0$. $\mathbf{E}_{t_1,t_2}$ is the joint effort decision $(E_{t_1},\dots,E_{t_2})$.

$X$: The latent mastery at position $t$. Discrete varaible. The cardinality of the possible state is $M_X$. $\mathbf{X}_{t_1,t_2}$ is the joint latent mastery $(X_{t_1},\dots,X_{t_2})$.

$Y$: The observed response at position $t$. Discrete variable. The cardinality of the possible state is $M_Y$. $\mathbf{Y}_{t_1,t_2}$ is the joint responses $(Y_{t_1},\dots,Y_{t_2})$.

$Z$: The learner's type. Dicrete variable. The cardinarlity of the possible learner type is $M_Z$

$\Omega$: The sample space of the re-parametrized joint distribution of observed response, stop decision and effort decision. Its cardinality is $N_{Omega}$.

$\alpha_z$: The probability that the learner is type Z.

$\beta_{z;k}$: The proportion hazard rate of one additional practice opportunity given latent mastery $k$ of learner type $z$ in the proportional hazard model

$c^{r,k}_j$: The probability of observing response $r$ given item $j$ and latent mastery level $k$.

$h_t^{z;k}$: The hazard rate at sequence $t$ given latent mastery $k$ of learner type $z$in the non-parametric hazard model.

$h_t^r$: The hazard rate at sequence $t$ given observed response $r$ in the non-parametric hazard model.

$\ell^{z;m,n}_{j}$: The pedagogical efficacy. The upper label describes the transition states ($m,n$) of the latent mastery for learner type $z$ while the lower label describes the tranisition conext,  item id ($j$).

$\pi^{z;k}$: The probability of latent mastery $k$ for learner type $z$ at the begining of the practice.

$\gamma_{j}^{z;k}$: The probability of exerting effort given item $j$ and latent mastery level $k$.

$\lambda_{z;k}$: The baseline hazard rate given latent mastery $k$ of learner type $z$ in the proportional hazard model






# Simulation Parameters

## The Monte Carlo Markov Chain Algorithm Convergence Demo: Single Type

The initial state density is:

$$
P(X)= \Bigg\{ \begin{array}{cc}
0.6 & \text{if }X=1 \\
0.3 & \text{if }X=2 \\
0.1 & \text{if }X = 3
\end{array}
$$
The state transtion matrix for item 1 is 

$$
\begin{bmatrix}
0.5 & 0.3 & 0.2\\
0 & 0.4 & 0.6\\
0 & 0 & 1
\end{bmatrix}
$$

The state transtion matrix for item 2 is 
$$
\begin{bmatrix}
0.2 & 0.3 & 0.5\\
0 & 0.7 & 0.3\\
0 & 0 & 1
\end{bmatrix}
$$

The observation matrix for item 1 is
$$
\begin{bmatrix}
0.8 & 0.2 & 0\\
0.2 & 0.6 & 0.2\\
0 & 0.2 & 0.8
\end{bmatrix}
$$

The observation matrix for item 2 is 
$$
\begin{bmatrix}
0.5 & 0.5 & 0\\
0.3 & 0.4 & 0.3\\
0 & 0.1 & 0.9
\end{bmatrix}
$$

The effort matrix for item 1 is
$$
\begin{bmatrix}
0.8 & 0.2\\
0.5 & 0.5\\
0.1 & 0.9
\end{bmatrix}
$$

The effort matrix for item 2 is 
$$
\begin{bmatrix}
0.7 & 0.3\\
0.4 & 0.6\\
0.01 & 0.99
\end{bmatrix}
$$

The baseline hazard rate is $\lambda=0.1$ for all states. 

The duration dependence is 

$$
\beta= \Bigg\{ \begin{array}{cc}
1.2 & \text{if }X=1 \\
1.1 & \text{if }X=2 \\
1 & \text{if }X = 3
\end{array}
$$


## The Monte Carlo Markov Chain Algorithm Convergence Demo: Multiple Types

The first type is 60% of the sample ($\alpha_1=0.6$). The first item occurs with 50% probability ($P(A_t=1)=0.5$).

For the type 1 learner, the initial mastery is 0.1 ($\pi_1=0.1$). The pedagogical efficacy of the first item is 0.3 ($\ell_{1,1}=0.3$), that of the second item is 0.7 ($\ell_{2,1}=0.7$).


For the type 1 learner, the initial mastery is 0.7 ($\pi_2=0.1$). The pedagogical efficacy of the first item is 0.2 ($\ell_{1,2}=0.2$), that of the second item is 0.8 ($\ell_{2,2}=0.8$).


For the both learner types, The observation matrix for item 1 is
$$
\begin{bmatrix}
0.9 & 0.1 \\
0.1 & 0.9
\end{bmatrix}
$$

The observation matrix for item 2 is 
$$
\begin{bmatrix}
0.8 & 0.2 \\
0.2 & 0.8 \\
\end{bmatrix}
$$

1000 learners are simulated. The practice sequence is 5.

## The Dynamic Selection Bias Demo

$$
\begin{aligned}
M_x&=2\\
M_y&=2\\
P(X_1=1) &= 0.4\\
P(X_t=1|X_{t-1}=0) &= 0.3\\
P(Y_t=1|X_t=1) &= 0.9\\
P(Y_t=1|X_t=0) &= 0.2\\
\end{aligned}
$$

## The Effort-induced Measurement Error Demo

$$
\begin{aligned}
P(X_0=1) = \pi &= 0.4\\
P(X_t=1|X_{t-1}=0,0)=\ell_0 &= 0.1\\
P(X_t=1|X_{t-1}=0,T)=\ell_T &= 0.7\\
P(X_t=1|X_{t-1}=0,C)=\ell_C &= 0.3\\
P(Y_1=1|X_t=1) =c_1^1&= 0.9\\
P(Y_1=1|X_t=0) =c_1^0&= 0.1\\
\end{aligned}
$$

In the effort choice model, the effort rate is specified to offset the difference in the pedagogical efficacy.All effort rates are 1 except for 

$$
\begin{aligned}
P(E_T=1|X_t=0) =e_T^0&= 0.3\\
P(E_C=1|X_t=0) =e_C^0&= 0.7\\
\end{aligned}
$$

3000 learners are simulated. Half of the learner receives item sequence (0,T,1) and the other receives item sequence (0,C,1).




# Answer Classification

The answers are initially classified into six categories:

(1) Blank answer: The learner submits nothing on the circumference and the area

(2) Non-blank wrong answer: Neither circumference nor area is correctly calculated and not includes in the slip or the wrong shape category 

(3) Slip: Both circumference and area are correctly calculated but the learner type in the wrong way. For example, switch the two blanks or add the unit of measurements

(4) Wrong Shape: The learner correctly calculates either the circumference or the area but for the small rectangle.

(5) wrong circumference: The learner correctly calculates the area of the large rectangle but not the circumference

(6) wrong area: The learner correctly calculates the circumference of the large rectangle but not the area

(7) Correct Answer: Both circumference and area of the large rectangle are correctly calculated

The Table  and Table  show the summary statistics of different groups' answer composition at a different stage of the experiment. There is a steady increase in the blank answer as the item sequence progresses. The nonblank answer increases most in the training session.


```{r, echo=FALSE,message=FALSE,warning=FALSE}
ans_composition = data  %>%
  group_by(group, qtype) %>% 
  summarize(blank = mean(blank_ans)*100,
            nonblank=mean(nonblank)*100,
            wrongshape=mean(wrong_shape)*100,
            rightcirc=mean(circ_right)*100,
            rightarea=mean(area_right)*100,
            slip=mean(is_slip)*100)
kable(
  ans_composition %>% arrange(qtype,group)%>% select(-rightcirc,-rightarea,-wrongshape,-slip), booktabs = TRUE,
  col.names=c('Group','Task','Blank Ans(%)','Non Blank Wrong Ans (%)'),
  align='c',
  caption = 'Answer Composition - All Wrong'
)
```

For learners who make a valid effort but still do not get all blanks correct, the percentage of the wrong shape declines steadily, but not the wrong area and the wrong circumference. In fact, there is a large increase in the wrong area in the post-test item and a large decrease in the wrong circumference. This observation is consistent with the step-wise efficacy analysis.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
kable(
  ans_composition %>% arrange(qtype,group) %>% select(-blank,-nonblank), booktabs = TRUE,
  col.names=c('Group','Task','Wrong Shape(%)', 'Wrong Area(%)', 'Wrong Circumference(%)','Slip(%)'),
  align='c',
  caption = 'Answer Composition - Partial Correct'
)
```

The reader may also be interested in the detailed breakdown of the answer composition. Other than the non-blank wrong answer, all four error categories have clustered answer patterns: The top 5 answers cover over 50% of the answers. The category of the non-blank answer has a wide dispersion. The top non-blank answer for all but the pre-test item is the correct answer to the pre-test question, which occupies about 30% of the non-blank answer.

As a reminder, the right answer for the pre-test item is (36,80), for the training item (28,48), for the post-test item (42,80). 


```{r, echo=FALSE,message=FALSE,warning=FALSE}
type_ans_stat = data %>% group_by(qtype, ans_type, raw_ans) %>% summarize(n=n()) %>%
  group_by(qtype,ans_type) %>% arrange(desc(n))


type_stat = data %>% group_by(qtype, ans_type) %>% summarize(N=n())

ans_stat = merge(type_ans_stat, type_stat) %>% mutate(pct=n/N) %>%
  filter(ans_type %in% c('wrong shape', 'slip','non-blank ans', 'right area','right circumference')) %>%
  group_by(qtype, ans_type) %>% arrange(qtype,  ans_type, desc(pct)) %>%
  mutate(cum_pct = cumsum(pct)) %>%
  mutate(idx = row_number()) %>% filter(idx<=5) %>%
 ungroup()  %>%
  select(qtype,ans_type, raw_ans,n, pct, cum_pct) 
```

For the non-blank answer, the most common error of the training item and the post-test item is (38,80), which is the correct answer to the pre-test item. The second common error for the post-test is (28,48), which is the correct answer to the training item.  The learners just memorized the answer key from the previous question, which signals that they are not making effort.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='non-blank ans')%>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Nonblank Answer'
)
```

The top category of the wrong shape is the right circumference and the right shape of the small rectangle, which is not surprising.

```{r,echo=FALSE,message=FALSE,warning=FALSE}


kable(
  ans_stat %>% filter(ans_type=='wrong shape') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Shape'
)
```

It is not surprising that either the circumference  or the double of the circumference of the small rectangle is in top 3 wrong answer category of the wrong circumference. In all three items, substitute the circumference with a (wrong) area is also a common mistake. For example, (160,80) in the pre-test and (24,48) in the training.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='right area') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Circumference'
)
```

The most common mistake of the wrong area is using the area of the small rectangle.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='right circumference') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Area'
)
```


Slip mainly involves adding 1 before or 0 after the right answer.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='slip') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Slip'
)
```
