# (APPENDIX) Appendix {-} 

# Notations

$i$: Index for learner. 

$j$: Index for practice item. The cardinality of the item id is $J$.

$t$: Index for practice sequence. The cardinality of the sequence id is $T$.

$A$: The assigment function that maps the items to the practice sequence. $A_t=j$ means assign item $j$ to sequence $t$. $\mathbf{A}_{t_1,t_2}$ is mapping from sequence position $t_1$ to $t_2$. 

$H$: The exit decision . Binary variable. If the learner stops at sequence $t$,$H_t=1$; otherwise, $H_t=0$. $\mathbf{H}_{t_1,t_2}$ is the joint exit decision $(H_{t_1},\dots,H_{t_2})$.

$E$: The effort decision. Binary variable. If the learner makes a legit effort at sequence $t$, $E_t=1$; otherwise, $E_t=0$. $\mathbf{E}_{t_1,t_2}$ is the joint effort decision $(E_{t_1},\dots,E_{t_2})$.

$X$: The latent mastery at position $t$. Discrete varaible. The cardinality of the possible state is $M_X$. $\mathbf{X}_{t_1,t_2}$ is the joint latent mastery $(X_{t_1},\dots,X_{t_2})$.

$Y$: The observed response at position $t$. Discrete varaible. The cardinality of the possible state is $M_Y$. $\mathbf{Y}_{t_1,t_2}$ is the joint responses $(Y_{t_1},\dots,Y_{t_2})$.


# Simulation Parameters of the Model Fit

The initial state density is:

$$
P(X)= \Bigg\{ \begin{array}{cc}
0.6 & \text{if }X=1 \\
0.3 & \text{if }X=2 \\
0.1 & \text{if }X = 3
\end{array}
$$
The state transtion matrix for item 1 is 

$$
\begin{bmatrix}
0.5 & 0.3 & 0.2\\
0 & 0.4 & 0.6\\
0 & 0 & 1
\end{bmatrix}
$$

The state transtion matrix for item 2 is 
$$
\begin{bmatrix}
0.2 & 0.3 & 0.5\\
0 & 0.7 & 0.3\\
0 & 0 & 1
\end{bmatrix}
$$

The observation matrix for item 1 is
$$
\begin{bmatrix}
0.8 & 0.2 & 0\\
0.2 & 0.6 & 0.2\\
0 & 0.2 & 0.8
\end{bmatrix}
$$

The observation matrix for item 2 is 
$$
\begin{bmatrix}
0.5 & 0.5 & 0\\
0.3 & 0.4 & 0.3\\
0 & 0.1 & 0.9
\end{bmatrix}
$$

The effort matrix for item 1 is
$$
\begin{bmatrix}
0.8 & 0.2\\
0.5 & 0.5\\
0.1 & 0.9
\end{bmatrix}
$$

The effort matrix for item 2 is 
$$
\begin{bmatrix}
0.7 & 0.3\\
0.4 & 0.6\\
0.01 & 0.99
\end{bmatrix}
$$

The baseline hazard rate is $\lambda=0.1$ for all states. 

The duration dependence is 

$$
\beta= \Bigg\{ \begin{array}{cc}
1.2 & \text{if }X=1 \\
1.1 & \text{if }X=2 \\
1 & \text{if }X = 3
\end{array}
$$

# Simulation Parameters of the BKT-Hazard Hybrid Model

$$
\begin{aligned}
M_x&=2\\
M_y&=2\\
P(X_1=1) &= 0.4\\
P(X_t=1|X_{t-1}=0) &= 0.3\\
P(Y_t=1|X_t=1) &= 0.9\\
P(Y_t=1|X_t=0) &= 0.2\\
\end{aligned}
$$


# Data Cleaning For Chapter 5
The study collected more than 68 million exercise logs. 

First retain serious learners, defined as those have more than 50 log entries, accounting for 20% of the total learners. However, serious learners generated 42 million exercises log or 62% of the total logs.

The recommendation algorithm is designed as much that the knowledge points alternate between different practice sequences so that the learner is not bored. Therefore, there are two sequence ranks one can compute. The global sequence rank ignores the interval and continues the practice rank count over the whole sample period. The local sequence rank only counts the practice rank within the sequence. Neither counting method is perfect. The global sequence rank completely ignores the dropout; while the local sequence rank violates the assumption of homogeneous initial mastery. For the purpose of this paper, learner heterogeneity is a lesser evil so the local sequence rank is chosen. 

If the learner has multiple practice sequences, choose the first one by calendar time.

The two knowledge points all have more than 200,000 practice logs. For two digit multiplication, 95% of the practice sequences have a life span smaller than 4 periods, 80% for vertical division. Because data are quite sparse past the fifth practice, the max practice sequence is 4. 

For the remaining valid data, take 40% random sample of training data and 20% random sample as out sample forecast. The analysis limits the size because the adaptive rejection sampling algorithm has trouble fitting large dataset. 



# Simulation Parameters of the Effort Induced Measurement Error

$$
\begin{aligned}
P(X_0=1) = \pi &= 0.4\\
P(X_t=1|X_{t-1}=0,0)=\ell_0 &= 0.1\\
P(X_t=1|X_{t-1}=0,T)=\ell_T &= 0.7\\
P(X_t=1|X_{t-1}=0,C)=\ell_C &= 0.3\\
P(Y_1=1|X_t=1) =c_1^1&= 0.9\\
P(Y_1=1|X_t=0) =c_1^0&= 0.1\\
\end{aligned}
$$

In the effort choice model, the effort rate is specified to offset the difference in the pedagogical efficacy.All effort rates are 1 except for 

$$
\begin{aligned}
P(E_T=1|X_t=0) =e_T^0&= 0.3\\
P(E_C=1|X_t=0) =e_C^0&= 0.7\\
\end{aligned}
$$

3000 learners are simulated. Half of the learner receives item sequence (0,T,1) and the other receives item sequence (0,C,1).



# The Log file
The log data collected for the experiment includes the following fields:

1. learner id
2. question id
3. Submit time
4. time spent on the question (seconds)
5. the grade (0-1)
6. the actual answer in the text

The time spent on the item is defined as the time elapsed between the time the server sends out the question to the learner's device and the time the learner's submitted answer on the last subquestion received by the server. The transmission time in the network is negligible, usually in the magnitude of 10 milliseconds. The time spent does not distinguish how much time spent on each sub question and it is not a clean measure of student's active learning time. What student did between the question  presentation and the answer submission is not observed.

Here is a sample of the original data log

```{r, echo=FALSE,message=FALSE,warning=FALSE}
kable(
  head(data%>%select(uid, eid, cmt_time, cmt_timelen,atag_pct, answers), 1), booktabs = TRUE,
  col.names = c('User ID', 'Item ID', 'Submit Time', 'Time Spent on Item', 'Score', 'Answers'),
  align='c',
  caption = 'Demo data from the Log'
)
```


# Answer Classification

The answers are initially classified into six categories:

(1) Blank answer: The learner submits nothing on the circumference and the area

(2) Non-blank wrong answer: Neither circumference nor area is correctly calculated and not includes in the slip or the wrong shape category 

(3) Slip: Both circumference and area are correctly calculated but the learner type in the wrong way. For example, switch the two blanks or add the unit of measurements

(4) Wrong Shape: The learner correctly calculates either the circumference or the area but for the small rectangle.

(5) wrong circumference: The learner correctly calculates the area of the large rectangle but not the circumference

(6) wrong area: The learner correctly calculates the circumference of the large rectangle but not the area

(7) Correct Answer: Both circumference and area of the large rectangle are correctly calculated

The Table  and Table  show the summary statistics of different groups' answer composition at a different stage of the experiment. There is a steady increase in the blank answer as the item sequence progresses. The nonblank answer increases most in the training session.


```{r, echo=FALSE,message=FALSE,warning=FALSE}
ans_composition = data  %>%
  group_by(group, qtype) %>% 
  summarize(blank = mean(blank_ans)*100,
            nonblank=mean(nonblank)*100,
            wrongshape=mean(wrong_shape)*100,
            rightcirc=mean(circ_right)*100,
            rightarea=mean(area_right)*100,
            slip=mean(is_slip)*100)
kable(
  ans_composition %>% arrange(qtype,group)%>% select(-rightcirc,-rightarea,-wrongshape,-slip), booktabs = TRUE,
  col.names=c('Group','Task','Blank Ans(%)','Non Blank Wrong Ans (%)'),
  align='c',
  caption = 'Answer Composition - All Wrong'
)
```

For learners who make a valid effort but still do not get all blanks correct, the percentage of the wrong shape declines steadily, but not the wrong area and the wrong circumference. In fact, there is a large increase in the wrong area in the post-test item and a large decrease in the wrong circumference. This observation is consistent with the step-wise efficacy analysis.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
kable(
  ans_composition %>% arrange(qtype,group) %>% select(-blank,-nonblank), booktabs = TRUE,
  col.names=c('Group','Task','Wrong Shape(%)', 'Wrong Area(%)', 'Wrong Circumference(%)','Slip(%)'),
  align='c',
  caption = 'Answer Composition - Partial Correct'
)
```

The reader may also be interested in the detailed breakdown of the answer composition. Other than the non-blank wrong answer, all four error categories have clustered answer patterns: The top 5 answers cover over 50% of the answers. The category of the non-blank answer has a wide dispersion. The top non-blank answer for all but the pre-test item is the correct answer to the pre-test question, which occupies about 30% of the non-blank answer.

As a reminder, the right answer for the pre-test item is (36,80), for the training item (28,48), for the post-test item (42,80). 


```{r, echo=FALSE,message=FALSE,warning=FALSE}
type_ans_stat = data %>% group_by(qtype, ans_type, raw_ans) %>% summarize(n=n()) %>%
  group_by(qtype,ans_type) %>% arrange(desc(n))


type_stat = data %>% group_by(qtype, ans_type) %>% summarize(N=n())

ans_stat = merge(type_ans_stat, type_stat) %>% mutate(pct=n/N) %>%
  filter(ans_type %in% c('wrong shape', 'slip','non-blank ans', 'right area','right circumference')) %>%
  group_by(qtype, ans_type) %>% arrange(qtype,  ans_type, desc(pct)) %>%
  mutate(cum_pct = cumsum(pct)) %>%
  mutate(idx = row_number()) %>% filter(idx<=5) %>%
 ungroup()  %>%
  select(qtype,ans_type, raw_ans,n, pct, cum_pct) 
```

For the non-blank answer, the most common error of the training item and the post-test item is (38,80), which is the correct answer to the pre-test item. The second common error for the post-test is (28,48), which is the correct answer to the training item.  The learners just memorized the answer key from the previous question, which signals that they are not making effort.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='non-blank ans')%>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Nonblank Answer'
)
```

The top category of the wrong shape is the right circumference and the right shape of the small rectangle, which is not surprising.

```{r,echo=FALSE,message=FALSE,warning=FALSE}


kable(
  ans_stat %>% filter(ans_type=='wrong shape') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Shape'
)
```

It is not surprising that either the circumference  or the double of the circumference of the small rectangle is in top 3 wrong answer category of the wrong circumference. In all three items, substitute the circumference with a (wrong) area is also a common mistake. For example, (160,80) in the pre-test and (24,48) in the training.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='right area') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Circumference'
)
```

The most common mistake of the wrong area is using the area of the small rectangle.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='right circumference') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Wrong Area'
)
```


Slip mainly involves adding 1 before or 0 after the right answer.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

kable(
  ans_stat %>% filter(ans_type=='slip') %>% select(-ans_type) , booktabs = TRUE,
  caption = 'Answer Breakdown: Slip'
)
```
